---
title: "Homework7"
author: "Amin Baabol"
date: "11/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Problem 5.12:


Instructions:

a) "Fit an ARIMA model"  -- specify the model and give the estimators of 
the coefficients. For example, if you use a AR(2) model to fit the data, then you 
need to give the model like y_t = 30 + 0.5 y_{t-1} + 0.3 y_{t-2} + \epsilon_t. 
Your answer should include the ACF and PACF of the data, then determine your 
"best" model. You can use the auto.arima() function to double check. 

"Investigate the model adequacy" --  show the 4-in-1 residual plots, and ACF and 
PACF plots for the residuals. Then interpret your outputs/results.

"Explain how this model would be used for forecasting" -- show the forecast model. 
To specify the forecast model, you need to show how to get estimators of the
$\Psi_i$'s. Show at least the first 4 $\Psi_i$'s. Similar to what we did for 
Example 5.3, 5.4, and 5.5 on page 381-382 in the class. The general forecast model 
is shown as the equation of (5.88) on page 379.

#Yt = Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t 
#Yt-mu = Î´ + Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t 

#Yt-mu = Î´ + Ï†1(yt-1) + Ï†2(yt-2) + ğœ€t 
#Yt = Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t
#yt=3396
#intercept = mean <- mu = 1545.4458
#Ï†1 = 1.1474, Ï†2 = -0.5997, yt-1 = 2657, yt-2 = 1590, ğœ€t = 601.838001
#yt <- ((1.1474*(2657-1545.4458)) - (0.5997*(1590-1545.4458)) + 601.838001)
#3396-1545.4458

### Part 5.2a

From the standard time series plot we can see cyclical pulse that doesn't have 
a fixed season.when you see this kind of undefined pattern, it is best to build 
an ARIMA model.The plot shows that when the mean reaches the peak it drops fairly
fast.There is not much middle ground.It goes from one extreme to the other which 
is an indicator of a possible autoregression.

The Augmented Dickey-Fuller Test indicates a statistically significant p-value 
of less then 0.01. Hence, we reject the null hypothesis that the time series is 
non-stationary in favor of the alternative hypothesis that the time series data 
is now stationary.Moreover,the ACF and PACF plots of the differenced time series 
data shows no significant initial lags,however, perhaps there seems to be a potential 
significant lag that crosses the threshold at around lag 12.Hence, It is wise to 
start with MA(0) then upgrade it to MA(1) or MA(2) to see if the MA component 
improves.The PACF plot also shows there could potentially be AR(0),AR(1),AR(2) 
or even AR(3).If the first couple of lags are significant we can be sure the model 
needs readjustment but the ACF plot shows latter lag 12 and PACT also shows latter 
lags 12,13 are significant which is a borderline situation because it could be 
by chance.So we are going to individual test the moving average component and the 
autoregressive components until we find the optimal AIC. Then we check the residuals 
using checkresiduals() from library(forecast) because there should be no more 
autoregression in them. The residuals should be random and normally distributed.


After individually tweaking the arima model's moving average component and the 
autoregressive component our manual parameter estimation indicates that we have 
two candidates for the q component. MA(0) yields an AIC of 302.2083 and MA(1) 
gives AIC of 302.4668. Furthermore, AR(0) or AR(1) for the p component respectively
give us AIC of 302.2083 and 302.6586.Since we performed a first order differencing 
to stabilize the mean our d parameter is going to be 1. Hence, our best models 
might be ARIMA(0,1,0) and ARIMA(1,1,1). Further comparison reveals that ARIMA model 
of the order (1,1,1) gives us the best criterion with the lowest AIC of 296.194.

After running model diagnostics to check model adequacy we see that the residuals
are normally distributed with a tiny hint of left skewness as shown in the histogram.
The qq-norm plot further supports that the normal distribution assumption isn't 
violated.The fitted vs. residuals indicate fairly random scatter centered around
0 with marginal difference of +/- 2. The variance seems to be slightly increasing 
as the fitted values increase.Additionally, the residual vs. observation order plot 
show that residuals exhibit normal random noise around 0 which suggests that there
is no serial correlation. The ACF of the residuals show a potential significance
at around lag 13 but after running a Ljung-Box's statistics we get a large p-value 
of 0.1654.This means we don't have enough statistical evidence to reject the null 
hypothesis which does not confirm the residual dependence or the residual autocorrelation
significance of lag 13. In essence, the apparent significant autocorrelation lag 
seen in the residuals ACF plot is due to some sort of random error and our 
ARIMA(1,1,1) is appropriate.


```{r}
#Raw data
Viscosity.Reading <- read.csv("~/Desktop/GradSchool/STATS 560 Time Series Analysis/Lecture/ViscosityData.csv",
                              header = TRUE)

#Convert the dataframe into time series
ViscosityReading.TS <- ts(Viscosity.Reading)

#Checking for NAs
summary(ViscosityReading.TS)

#time series plot
library(lattice)
xyplot.ts(ViscosityReading.TS,
          main = 'Figure 1: Time Series Plot of TABLE B.3 Data',
          xlab = 'Time Period',
          ylab = 'Viscosity Reading')


#Checking for stationarity
library(tseries)
library(forecast)
library(zoo)


#Stationarity check:ACF and PACF plots
tsdisplay(ViscosityReading.TS)

#Stationarity:Dicke-Fuller test 
adf.test(ViscosityReading.TS)

#The sample ACF plot suggests MA(2), MA(3) or even MA(4) models.
#similarly, it be interpreted as an exponential decay pattern suggesting
#AR(p) models. Then we check sample PACF plot and conclude it cuts 
#after 1 suggesting AR(1) model.


AR.first <- Arima(ViscosityReading.TS, order = c(1,0,0))

AR.first

#Series: ViscosityReading.TS 
#ARIMA(1,0,0) with non-zero mean 

#Coefficients:
         #ar1     mean
      #0.7859  84.9827
#s.e.  0.0605   0.4649

#sigma^2 estimated as 1.082:  log likelihood=-145.32
#AIC=296.64   AICc=296.89   BIC=304.45

#Show in New WindowClear OutputExpand/Collapse Output

#Call:
#arima(x = ViscosityReading.TS, order = c(1, 0, 0))

#Coefficients:
         #ar1  intercept
      #0.7859    84.9827
#s.e.  0.0605     0.4649

#sigma^2 estimated as 1.061:  log likelihood = -145.32,  aic = 296.64

#Yt = Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t 
#Yt-mu = Î´ + Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t 

#Yt-mu = Î´ + Ï†1(yt-1) + Ï†2(yt-2) + ğœ€t 
#Yt = Ï†1(yt-1 - mu) + Ï†2(yt-2 - mu) + ğœ€t

#Î´ = 84.9827*(1-)

#First order differencing to obtain ARIMA parameter d
ViscosityReading.Diff <- diff(ViscosityReading.TS, differences = 1)


#checking stationarity of the difference time series

#Stationarity check:ACF and PACF plots
tsdisplay(ViscosityReading.Diff)

#Stationarity:Dickey-Fuller test 
adf.test(ViscosityReading.Diff)


#Comparing AIC performance 
MA.0 <- Arima(ViscosityReading.Diff, order = c(0,0,0))
MA.1 <- Arima(ViscosityReading.Diff, order = c(0,0,1))
MA.2 <- Arima(ViscosityReading.Diff, order = c(0,0,2))
MA.3 <- Arima(ViscosityReading.Diff, order = c(0,0,3))

AR.0 <- Arima(ViscosityReading.Diff, order = c(0,0,0))
AR.1 <- Arima(ViscosityReading.Diff, order = c(1,0,0))
AR.2 <- Arima(ViscosityReading.Diff, order = c(2,0,0))
AR.3 <- Arima(ViscosityReading.Diff, order = c(3,0,0))

ARIMA.010 <- Arima(ViscosityReading.TS, order = c(0,1,0))
ARIMA.111 <- Arima(ViscosityReading.TS, order = c(1,1,1))



paste("MA(0) AIC")
AIC(MA.0)
paste("MA(1) AIC")
AIC(MA.1)
paste("MA(2) AIC")
AIC(MA.2)
paste("MA(3) AIC")
AIC(MA.3)


paste("AR(0) AIC")
AIC(AR.0)
paste("AR(1) AIC")
AIC(AR.1)
paste("AR(2) AIC")
AIC(AR.2)
paste("AR(3) AIC")
AIC(AR.3)


paste("ARIMA(0,1,0) AIC")
AIC(ARIMA.010)
paste("ARIMA(1,1,1) AIC")
AIC(ARIMA.111)


#Checking the residuals of ARIMA(1,1,1) model
checkresiduals(ARIMA.111)


#Another method to run model diagnostics(from the book)
ARIMA111.Fitted<-as.vector(fitted(ARIMA.111))
ARIMA111.Residuals <-as.vector(residuals(ARIMA.111))


Box.test(ARIMA111.Residuals,
         lag = 20,
         fitdf = 18,
         type = "Ljung")

#4-in-1 plot of the residuals
par(mfrow = c(2,2),oma = c(0,0,0,0))
qqnorm(ARIMA111.Residuals,
       datax = TRUE,
       pch = 16,
       xlab = 'Residual',
       main = '')
qqline(ARIMA111.Residuals,
       datax = TRUE)
plot(ARIMA111.Fitted,
     ARIMA111.Residuals,
     pch = 16,
     xlab = 'Fitted Value',
     ylab = 'Residual')
abline(h = 0)
hist(ARIMA111.Residuals,
     col = "gray",
     xlab = 'Residual',
     main = '')
plot(ARIMA111.Residuals,
     type = "l",
     xlab = 'Observation Order',
     ylab = 'Residual')
points(ARIMA111.Residuals,
       pch = 16,
       cex = .5)
abline(h = 0)


#check against auto.arima()
auto.arima(TableB3.difference,
           trace = TRUE,
           stepwise = FALSE,
           approximation = FALSE)


```





b. "Forecast the last 20 observations." -- No matter what model that you got in 
part a, to answer question b, please use auto.arima() to get the best model. 
(This is just for the grading purpose.) Then use the forecast() to get the 
forecast values. See the R code on page 408. This will give you a 1- to 20- step 
ahead forecasts.

c. "Show how to obtain prediction intervals..." --- The 80% and 95% Prediction 
Intervals should have been obtained from part b if you use forecast() function. 
You don't have to  calculate the Prediction intervals again. For this question, 
just show the prediction interval formula. 

```{r }
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
